A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-01-23 paper/paper.tex
some clarification

\subsection{Working with sets in type theory}
Martin-Löf's Type theory~\cite{martinlof75,martinlof84} is a logical formalism based on the notion of a \emph{type} rather than that of a\emph{set}. It can be seen as a foundation of mathematics alternative to set theory and is the core of several tools for the formalization of mathematics such as Coq~\cite{coq22} or Lean~\cite{lean15}. Type theory is a flexible formalism supporting different models. One of this model interprets types as topological spaces, where equality is interpreted as path and substitutivity of equality as transport. This model supports the univalence principle stating that equality of types mimics equivalence of types, leading to the development of Homotopy Type Theory~\cite{hottbook}. In type theory, types are organized in a hierarchy of universes written $\Type_m$ for $m$ a natural number. Main types in type theory are the types of dependent pairs, written $\Sigma a : A.\,B(a)$, and the types of dependent functions, written $\Pi a:A.\,B(a)$, for $A$ a type and $B(a)$ a type dependent on the inhabitant $a$ of $A$. We assume our type theory to also include a distinguished singleton type, written $\unittype$, and with inhabitant $\unitpoint$, the type of boolean values, called $\textsf{bool}$, and the type of natural numbers. We also write $\hd$ and $\tl$ the projections of dependent pairs. In type theory, logical propositions are types themselves and we use $\Pi$ to represent universal quantification and $\Sigma$ to represent existential quantification.

@@

Most components of the construction takes inequality constraints as parameters, and we have left implicit that they are satisfied in the tables. Comparison of inequality proofs depends on the definition of inequality, which we leave implicit, and assume that we can pick a definition for which proofs of $m \leq n$ are provably unique and thus definitionally equal by the reflection rule.

%TODO add a word on the definition of eq on sig and pi, about eq_trans, etc.

Remark: For a fixed constant $n$, the coherence conditions evaluate to a reflexivity proof, so that the construction evaluates to an effective sequence of types of iterated relations not mentioning $\restrf$ nor $\cohf$ anymore.

@@

Since we have formalized the proof in the Coq proof assistant, which works in intensional type theory, we now present the tables in \emph{intensional} type theory, leaving implicit some details, which we will explain here, and other details, which we will explain in \ref{sec:mech}.

Tables \ref{fulltab:coind}, \ref{fulltab:core}, and \ref{fulltab:frames} are unchanged, except that we make explicit all arguments, including those that can be inferred from the context. The need for transport along a proof of commutation of $\restrf$ in the definition of $\restrl$ is made explicit as shown in table \ref{fulltab:faces}, where the arrow over $\cohf$ indicates the direction of rewrite. The proof of $\cohf$ is described on \ref{fulltab:coh}, and it requires explication of several equalities which were implicit in extensional type theory. The commutation of $\restrl$ lives in a type referring to $\cohf$, so we need a transport along the commutation of $\restrf$ in the statement of $\cohl$. The proof of $\cohl$ is the most involved proof in the construction. This is where the higher-dimensional coherence condition is needed. The exact formulation of this coherence condition is as follows.

>>>>

\subsection{Working with sets in type theory}
Martin-Löf's Type theory~\cite{martinlof75,martinlof84} is a logical formalism based on the notion of a \emph{type} rather than that of a \emph{set}. It can be seen as a foundation of mathematics alternative to set theory and is the core of several tools for the formalization of mathematics such as Coq~\cite{coq22} or Lean~\cite{lean15}. In type theory, propositions are types and proofs are programs. A particularity of type theory is also that types and programs, henceforth propositions and proofs also, are considered modulo an equational theory called definitional equality.

Type theory is a flexible formalism supporting different models. One of this model interprets types as topological spaces, where equality is interpreted as path and substitutivity of equality as transport. This model supports the univalence principle stating that equality of types mimics equivalence of types, leading to the development of Homotopy Type Theory~\cite{hottbook}. In type theory, types are organized in a hierarchy of universes written $\Type_m$ for $m$ a natural number. Main types in type theory are the types of dependent pairs, written $\Sigma a : A.\,B(a)$, the types of dependent functions, written $\Pi a:A.\,B(a)$, for $A$ a type and $B(a)$ a type dependent on the inhabitant $a$ of $A$, and the type of propositional equality, written $t = u$. We assume our type theory to also include a distinguished singleton type, written $\unittype$, and with inhabitant $\unitpoint$, the type of boolean values, called $\textsf{bool}$, and the type of natural numbers. We also write $\hd$ and $\tl$ the projections of dependent pairs. In type theory, logical propositions are types themselves and we use $\Pi$ to represent universal quantification and $\Sigma$ to represent existential quantification.

@@

Most components of the construction takes inequality constraints as parameters, and we have left implicit that they are satisfied in the tables. Comparison of inequality proofs depends on the definition of inequality, which we leave implicit, only assuming that we can pick a definition for which proofs of $m \leq n$ are provably unique and thus definitionally equal by the reflection rule.

The construction takes benefit of various provable equalities over proofs of equality being definitional by the reflection rule. This includes in particular the groupoid properties of equality. Notably, unicity of identity proofs holds in extensional type theory, so that any type is automatically an $\U$. Also, we left implicit in table~\ref{tab:coh} the use of the isomorphism between $u = v$ and $\Sigma (p:u.\hd = v.\hd). (u.\tl = v.\tl)$ for $u$ and $v$ in a $\Sigma$-type. In the same table, we also left implicit the use of the isomorphism between $f = g$ and $\Pi a:A.f(a) = g(a)$ for $f$ and $g$ in $\Pi a:A.\, B$, where it should be recalled that the right-to-left map, that is function extensionality, holds by default in extensional type theory.

% TODO explain the role of the subscript HSet on =

As a side remark, note that for a fixed constant $n$, the coherence conditions evaluate to a reflexivity proof, so that the construction evaluates to an effective sequence of types of iterated relations not mentioning $\restrf$ nor $\cohf$ anymore.

@@

Since we have formalized the proof in the Coq proof assistant, which implements an intensional type theory, we now present the tables in \emph{intensional} type theory~\cite{martin-lof...,Coq}, leaving implicit some details, which we will explain here, and other details, which we will explain in \ref{sec:mech}.

Tables \ref{fulltab:coind}, \ref{fulltab:core}, and \ref{fulltab:frames} are unchanged, except that we make explicit all arguments, including those that can be inferred from the context. The need for transport along a proof of commutation of $\restrf$ in the definition of $\restrl$ is made explicit as shown in table \ref{fulltab:faces}, where the arrow over $\cohf$ indicates the direction of rewrite. The proof of $\cohf$ is described on \ref{fulltab:coh}, and it requires explicitation of several equalities which were definitional in extensional type theory. The commutation of $\restrl$ lives in a type referring to $\cohf$, so we need a transport along the commutation of $\restrf$ in the statement of $\cohl$. The proof of $\cohl$ is the most involved proof in the construction. This is where the higher-dimensional coherence condition is needed. The exact formulation of this coherence condition is as follows.
